---
title: "Subset Large VCF File"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{create_test_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mpbr)
```

# Introduction

The functions in {mpbr} are intended to extract and build SNPs data for the
downstream population genetic analyses. With the increase in the performance of
the sequencing platforms, the variant callers result in large VCF files with a
huge amount of variants. This vignette describes the process of subsetting
a large VCF file to generate a subset dataset that both users and maintainers
can use to test the package functionalities.

## Test data creation process

The subsetting process accounts for the size of the different chromosomes. A 100
sites will be randomly selected from the widest chromosome. The number of
selected sites from the remaining chromosomes will be based on the ration
between the length of a given chromosome and the widest chromosome.

```{r}
chroms <- c("Pf3D7_01_v3", "Pf3D7_02_v3", "Pf3D7_03_v3", "Pf3D7_04_v3",
            "Pf3D7_05_v3", "Pf3D7_06_v3", "Pf3D7_07_v3", "Pf3D7_08_v3",
            "Pf3D7_09_v3", "Pf3D7_10_v3", "Pf3D7_11_v3", "Pf3D7_12_v3",
            "Pf3D7_13_v3", "Pf3D7_14_v3")
chrom_length <- c(640851, 947102, 1067971, 1200490, 1343557, 1418242, 1445207,
                  1472805, 1541735, 1687656, 2038340, 2271494, 2925236, 3291936)
subset_sizes <- ceiling((chrom_length / max(chrom_length)) * 100)
names(subset_sizes) <- chroms
print(subset_sizes)
```

The `r system.file("scripts", "subset_large_vcf.sh", package = "mpbr")` file
contains the bash script built to perform the subsetting of the initial VCF
file.






