---
title: "design-decision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{design-decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mpbr)
```

## Mixed genotypes Phasing

Phasing is the process of transforming a mixed allele into either the reference
or alternate allele. There are several statistical approaches that can be used
to perform mixed genotype phasing. The following method is implemented in the `phase()` function:

* All mixed genotypes will be transformed into `0` or `1` based on the allele
with the highest supporting reads at the corresponding positions, except when
`heterozygous` is set to `TRUE`. When this option is activated, the function
will identify and transform the `true` mixed alleles into `2`. An allele is
considered as a `true` mixed allele when the following two conditions are met:
  * the `ratio` in allelic depth is greater than the specified ratio cut-off,
  * and the `min_ad` (minimum allelic depth between the reference and alternate
  alleles) is greater than or equal to the specified minimum allelic depth
  cut-off.
* When the two alleles have an equal allelic depth (AD), then the mixed genotype
is transformed based on the value of the `phase_equal_ad` argument. Its possible
values are:
  * `bernoulli`: the mixed allele is recoded into `0` or `1` based on a
  Bernoulli simulation with a parameter equal to the minor allele frequency of
  the SNPs on which the site is found.
  * `minor_allele`: the mixed allele is transformed into the least represented
  allele (i.e. the allele with the minor frequency in the population).

The `phase()` function takes the following arguments:

* `snpdata`: the input <SNPdata> object
* `genotype`: the name of the genotype matrix that contains the mixed alleles to
be phased.
* `heterozygous`: a logical that is used to determine whether keep `true`
heterozygous sites (recode them as `2`) or not. Default is `FALSE`.
* `ration`: a numeric that varies between 0 and 1. It represents the ratio in
allelic depth cut-off. Default is 0.1.
* `min_ad`: an integer that represents the minimum allelic depth cut-off.
Default is 2. A site is considered as a `true` heterozygous site when the
`ratio` and the `min_ad` criteria are satisfied.
* `ncores`: an integer that represents the number of cores to be used during the
phasing process. Default is 4.

